---
title: "EdgeR normalization of bowtie2 mapping to consenus MAGs post aggregation"
output: html_notebook
---

Normalize each of three time series individually and combined using the time series
splines fit method recommened by edgeR. For the filtering step (filterByExpr), all edgeR defaults were used except
min.count=1 (at least 1 count per group) Reads were mapped using bowtie2, filtered for quality,
then counted using htseq-count. Here we sum the gene clusters for each lineage prior to normalization with edgeR


```{r}
library(tidyverse)
library(edgeR)
library(limma)
library(statmod)
library(splines)
library(data.table)
```



```{r}
#import htseq-count output
#all counts are in a single file
htdf = read_csv("2022.09.22_htseqCount_dataframe_backup.csv")

```


```{r}
#groupby lineage, gene cluster and sum
gcdf = htdf %>% group_by(Lineage, `Gene Cluster`) %>%summarize_at(vars(starts_with("Metofsampl")),sum)
```


```{r}
#merge columns to create a unique ID for the rowname (format required by edgeR)
gcdf = gcdf %>% unite("uid",Lineage:`Gene Cluster`, sep="_")%>% column_to_rownames("uid")
```

```{r}
#export gcdf
write.csv(gcdf,"GeneClusters_rawcounts.csv")
```


```{r}
#define time series metadata
ms2005names =c("MetofsampleMSt1","MetofsampleMSt2","MetofsampleMSt3","MetofsampleMSt4",
               "MetofsampleMSt5","MetofsampleMSt6","MetofsampleMSt7","MetofsampleMSt8",
               "MetofsampleMSt9","MetofsamplMSP12","MetofsamplMSt11","MetofsamplMSt12")
ms2005times = c(0, 120, 240,360,480,600,720,840,960,1080,1200,1320)#starting from first sample=0
ms2005groups = c(1,2,3,4,5,6,7,8,9,10,11,12)

ms2009names = c("MetofsampleR2cd","MetofsampleR3cd","MetofsampleR5cd","MetofsampleR6cd",
                "MetofsampleR7cd","MetofsampleR8cd","MetofsampleR9cd","MetofsamplR10cd",
                "MetofsamplR11cd","MetofsamplR12cd","MetofsamplR13cd","MetofsamplR14cd")
ms2009times = c(0, 40, 165, 275, 575, 650, 735, 795, 882,1015, 1120, 1235)
ms2009groups = c(1,2,3,4,5,6,7,8,9,10,11,12)

os2009names = c("Metofsamplet1cd","Metofsamplet2cd", "Metofsamplet4cd","Metofsamplet5cd",
                "Metofsamplet8cd","Metofsamplt11cd", "Metofsamplt13cd", "Metofsamplt14cd")
os2009times = c(0, 105,195,240,780,915,1080,1200)
os2009groups = c(1,2,3,4,5,6,7,8)

allsamples = c("MetofsampleMSt1","MetofsampleMSt2","MetofsampleMSt3","MetofsampleMSt4",
               "MetofsampleMSt5","MetofsampleMSt6","MetofsampleMSt7","MetofsampleMSt8",
               "MetofsampleMSt9","MetofsamplMSP12","MetofsamplMSt11","MetofsamplMSt12",
             "Metofsamplet1cd","Metofsamplet2cd", "Metofsamplet4cd","Metofsamplet5cd",
                "Metofsamplet8cd","Metofsamplt11cd", "Metofsamplt13cd", "Metofsamplt14cd",
             "MetofsampleR2cd","MetofsampleR3cd","MetofsampleR5cd","MetofsampleR6cd",
                "MetofsampleR7cd","MetofsampleR8cd","MetofsampleR9cd","MetofsamplR10cd",
                "MetofsamplR11cd","MetofsamplR12cd","MetofsamplR13cd","MetofsamplR14cd") #artificially concatenated ms2005, os2009, ms2009 - first sample = 0
alltimes = c(0, 120, 240,360,480,600,720,840,960,1080,1200,1320,1320,1425,1515,
             1560,2100,2235,2400,2520,2530,2570,2695,2805,3105,3180,3265,3325,
             3412,3545,3650,3765)
allgroups = c(1,2,3,4,5,6,7,8,9,10,11,12,13,14,15,16,17,18,19,20,21,22,23,24,25,26,
              27,28,29,30,31,32)
```

```{r}
#separate hdtf into each time series and order columns if needed
ms2005df = gcdf[,ms2005names]
ms2009df = gcdf[,ms2009names]
os2009df = gcdf[,os2009names]
orderedgcdf = gcdf[,allsamples]
```

From here, run edgeR with the following options: RLE. natural spline with 3 degree of freedom 
on each time series and the combined dataframe (minus R15cd which lacks metadata).

MS2005
```{r}
#input into the DE analysis pipeline
yms2005 <- DGEList(counts=ms2005df,group=ms2005groups)

#remove genes that fit the default parameters for low expression in edgeR
keepms2005 <-filterByExpr(yms2005, min.count=1)
table(keepms2005)

#use the library size calculation in edgeR with the RLE option
yms2005 <- yms2005[keepms2005, , keep.lib.sizes=FALSE]
yms2005 <- calcNormFactors(yms2005, method="RLE")
yms2005$samples
```
```{r}
#model as a 3-degree of freedom spline fit
Xms2005 <- ns(ms2005times, df=3)
designms2005 <- model.matrix(~ Xms2005)
designms2005
#estimate the dispersion
yms2005 <- estimateDisp(yms2005, designms2005)
sqrt(yms2005$common.dispersion)
#plot the dispersion
plotBCV(yms2005)
#this estimates the coeffcients for each gene
fitms2005 <- glmQLFit(yms2005, designms2005, robust=TRUE)
plotQLDisp(fitms2005)

#this looks at the fit of the model across the genes
fitms2005 <- glmQLFTest(fitms2005, coef=2:4)

#use the default significance test
summary(decideTests(fitms2005))
```
```{r}
tabms2005 <- as.data.frame(topTags(fitms2005, n=nrow(ms2005df)))
```

```{r}
#export the results
#generate the line fits for each gene
logCPM.obs.ms2005 <- cpm(yms2005, normalized.lib.sizes = TRUE, log=TRUE, prior.count=fitms2005$prior.count)
CPM.obs.ms2005 <-  cpm(yms2005, normalized.lib.sizes = TRUE, log=FALSE, prior.count=fitms2005$prior.count)
logCPM.fit.ms2005 <- cpm(fitms2005, normalized.lib.sizes = TRUE, log=TRUE)
```
```{r}
#export the RLE normalization, CPM, and log-CPM to files
write.csv(yms2005[["samples"]],"Geneclusters_mincount1_MS2005_consensusMAGs_bowtie_library_normalization.csv")
write.csv(CPM.obs.ms2005,"Geneclusters_mincount1_MS2005_consensusMAGs_bowtie_normalized_CPM.csv")
write.csv(logCPM.obs.ms2005,"Geneclusters_mincount1_MS2005_consensusMAGs_bowtie_normalized_log2CPM.csv")
write.csv(tabms2005,"Geneclusters_mincount1_MS2005_consensusMAGs_bowtie_edgeR_RLE_ns_fit.csv")
```
MS2009

```{r}
#input into the DE analysis pipeline
yms2009 <- DGEList(counts=ms2009df,group=ms2009groups)

#remove genes that fit the default parameters for low expression in edgeR
keepms2009 <-filterByExpr(yms2009, min.count=1)
table(keepms2009)

#use the library size calculation in edgeR with the RLE option
yms2009 <- yms2009[keepms2009, , keep.lib.sizes=FALSE]
yms2009 <- calcNormFactors(yms2009, method="RLE")
yms2009$samples
```
```{r}
#model as a 3-degree of freedom spline fit
Xms2009 <- ns(ms2009times, df=3)
designms2009 <- model.matrix(~ Xms2009)
designms2009
#estimate the dispersion
yms2009 <- estimateDisp(yms2009, designms2009)
sqrt(yms2009$common.dispersion)
#plot the dispersion
plotBCV(yms2009)
#this estimates the coeffcients for each gene
fitms2009 <- glmQLFit(yms2009, designms2009, robust=TRUE)
plotQLDisp(fitms2009)

#this looks at the fit of the model across the genes
fitms2009 <- glmQLFTest(fitms2009, coef=2:4)

#use the default significance test
summary(decideTests(fitms2009))
```
```{r}
tabms2009 <- as.data.frame(topTags(fitms2009, n=nrow(ms2009df)))
```

```{r}
#export the results
#generate the line fits for each gene
logCPM.obs.ms2009 <- cpm(yms2009, normalized.lib.sizes = TRUE, log=TRUE, prior.count=fitms2009$prior.count)
CPM.obs.ms2009 <-  cpm(yms2009, normalized.lib.sizes = TRUE, log=FALSE, prior.count=fitms2009$prior.count)
logCPM.fit.ms2009 <- cpm(fitms2009, normalized.lib.sizes = TRUE, log=TRUE)
```
```{r}
#export the RLE normalization, CPM, and log-CPM to files
write.csv(yms2009[["samples"]],"Geneclusters_mincount1_MS2009_consensusMAGs_bowtie_library_normalization.csv")
write.csv(CPM.obs.ms2009,"Geneclusters_mincount1_MS2009_consensusMAGs_bowtie_normalized_CPM.csv")
write.csv(logCPM.obs.ms2009,"Geneclusters_mincount1_MS2009_consensusMAGs_bowtie_normalized_log2CPM.csv")
write.csv(tabms2009,"Geneclusters_mincount1_MS2009_consensusMAGs_bowtie_edgeR_RLE_ns_fit.csv")
```
OS2009

```{r}
#input into the DE analysis pipeline
yos2009 <- DGEList(counts=os2009df,group=os2009groups)

#remove genes that fit the default parameters for low expression in edgeR
keepos2009 <-filterByExpr(yos2009, min.count=1)
table(keepos2009)

#use the library size calculation in edgeR with the RLE option
yos2009 <- yos2009[keepos2009, , keep.lib.sizes=FALSE]
yos2009 <- calcNormFactors(yos2009, method="RLE")
yos2009$samples
```
```{r}
#model as a 3-degree of freedom spline fit
Xos2009 <- ns(os2009times, df=3)
designos2009 <- model.matrix(~ Xos2009)
designos2009
#estimate the dispersion
yos2009 <- estimateDisp(yos2009, designos2009)
sqrt(yos2009$common.dispersion)
#plot the dispersion
plotBCV(yos2009)
#this estimates the coeffcients for each gene
fitos2009 <- glmQLFit(yos2009, designos2009, robust=TRUE)
plotQLDisp(fitos2009)

#this looks at the fit of the model across the genes
fitos2009 <- glmQLFTest(fitos2009, coef=2:4)

#use the default significance test
summary(decideTests(fitos2009))
```
```{r}
tabos2009 <- as.data.frame(topTags(fitos2009, n=nrow(os2009df)))
```

```{r}
#export the results
#generate the line fits for each gene
logCPM.obs.os2009 <- cpm(yos2009, normalized.lib.sizes = TRUE, log=TRUE, prior.count=fitos2009$prior.count)
CPM.obs.os2009 <-  cpm(yos2009, normalized.lib.sizes = TRUE, log=FALSE, prior.count=fitos2009$prior.count)
logCPM.fit.os2009 <- cpm(fitos2009, normalized.lib.sizes = TRUE, log=TRUE)
```
```{r}
#export the RLE normalization, CPM, and log-CPM to files
write.csv(yos2009[["samples"]],"Geneclusters_mincount1_OS2009_consensusMAGs_bowtie_library_normalization.csv")
write.csv(CPM.obs.os2009,"Geneclusters_mincount1_OS2009_consensusMAGs_bowtie_normalized_CPM.csv")
write.csv(logCPM.obs.os2009,"Geneclusters_mincount1_OS2009_consensusMAGs_bowtie_normalized_log2CPM.csv")
write.csv(tabos2009,"Geneclusters_OS2009_mincount1_consensusMAGs_bowtie_edgeR_RLE_ns_fit.csv")
```
Concatenated time series

```{r}
#input into the DE analysis pipeline
yall <- DGEList(counts=orderedgcdf,group=allsamples)

#remove genes that fit the default parameters for low expression in edgeR
keepall <-filterByExpr(yall, min.count=1)
table(keepall)

#use the library size calculation in edgeR with the RLE option
yall <- yall[keepall, , keep.lib.sizes=FALSE]
yall <- calcNormFactors(yall, method="RLE")
yall$samples
```
```{r}
#model as a 3-degree of freedom spline fit
Xall <- ns(alltimes, df=3)
designall <- model.matrix(~ Xall)
designall
#estimate the dispersion
yall <- estimateDisp(yall, designall)
sqrt(yall$common.dispersion)
#plot the dispersion
plotBCV(yall)
#this estimates the coeffcients for each gene
fitall <- glmQLFit(yall, designall, robust=TRUE)
plotQLDisp(fitall)

#this looks at the fit of the model across the genes
fitall <- glmQLFTest(fitall, coef=2:4)

#use the default significance test
summary(decideTests(fitall))
```
```{r}
taball <- as.data.frame(topTags(fitall, n=nrow(orderedgcdf)))
```

```{r}
#export the results
#generate the line fits for each gene
logCPM.obs.all <- cpm(yall, normalized.lib.sizes = TRUE, log=TRUE, prior.count=fitall$prior.count)
CPM.obs.all <-  cpm(yall, normalized.lib.sizes = TRUE, log=FALSE, prior.count=fitall$prior.count)
logCPM.fit.all <- cpm(fitall, normalized.lib.sizes = TRUE, log=TRUE)
```
```{r}
#export the RLE normalization, CPM, and log-CPM to files
write.csv(yall[["samples"]],"Geneclusters_mincount1_all_consensusMAGs_bowtie_library_normalization.csv")
write.csv(CPM.obs.all,"Geneclusters_mincount1_all_consensusMAGs_bowtie_normalized_CPM.csv")
write.csv(logCPM.obs.all,"Geneclusters_mincount1_all_consensusMAGs_bowtie_normalized_log2CPM.csv")
write.csv(taball,"Geneclusters_mincount1_all_consensusMAGs_bowtie_edgeR_RLE_ns_fit.csv")
```


